---
title: "A correlation analysis of clinical variables of TCGA-KIRC patients"
output: 
  html_document: 
    default
  github_document: 
    df_print: paged
    html_preview: FALSE
    keep_html: TRUE
  pdf_document:
    latex_engine: xelatex
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all") })     
---

This project contains a pipeline of clinical analysis of the Cancer Genome Atlas Kidney Renal Clear Cell Carcinoma (TCGA-KIRC) data of patients, from [Genomic Data Commons Data Portal](https://portal.gdc.cancer.gov/exploration?filters=%7B%22op%22%3A%22and%22%2C%22content%22%3A%5B%7B%22op%22%3A%22in%22%2C%22content%22%3A%7B%22field%22%3A%22cases.project.project_id%22%2C%22value%22%3A%5B%22TCGA-KIRC%22%5D%7D%7D%5D%7D) and [cBioPortal](https://www.cbioportal.org/study/summary?id=kirp_tcga).

Previously, we presented [an exploratory preprocessing analysis](1.preprocessing.md). In this section, Chi-squared test is applied to compare two or more proportions of categorical variables and T-student test to compare the means of numeric ones regardind the levels of 'Overall_Survival_Status'. The Hypoteis test is performed and p-value indicates the strength of evidence in supportting the null hypothesis.


```{r error=TRUE, message=FALSE, warning=FALSE, include=FALSE, purl=FALSE, results='hide'}
## This chunk automatically generates a text .R version of this script when running within knitr.
input  = knitr::current_input()  # filename of input document
output = paste(tools::file_path_sans_ext(input), 'R', sep = '.')
knitr::purl(input,output,documentation=2,quiet=T)
# Avoid duplicate label error of knitr::purl
options(knitr.duplicate.label = 'allow')
# Code to browse the markdown file with renderized images.
knitr::opts_chunk$set(
  fig.path = "figs/render-"
)
```


```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Set the packages of interest
packages = c("tidyverse","skimr","finalfit","rstatix","ggpubr","GGally")

# if a package is installed, it will be loaded
# otherwise, the missing package(s) will be installed and loaded
package.check <- lapply(packages, FUN = function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x, dependencies = TRUE)
    library(x, character.only = TRUE)
  }
})

suppressMessages(library("tidyverse"))
setwd(".")
```


## 1. Importing data

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

kirc_clinic <- read_csv("data/kirc_clinic.csv")

```

## 2. Taming data 

```{r}
kirc_clinic <- kirc_clinic %>%
  mutate_if(is.character, as.factor) %>%
  mutate(patient_id = as.character(patient_id))
```

## 3. Checking categorical variables

check frequency, lables and levels 

```{r echo=FALSE}

kirc_clinic %>%
  select_if(is.factor) %>%
  summary() 
```

## 4. Checking variables

```{r}
glimpse(kirc_clinic)
skim(kirc_clinic) 
#View(kirc_clinic)
```

## 5. Numeric variables vs. over_surv_stt
graphic visualization and t-test


```{r message=FALSE, warning=FALSE}

cols_numeric <- kirc_clinic %>% select_if(is.numeric) %>% names

kirc_clinic_numeric <- kirc_clinic %>%
                      select(one_of(c(cols_numeric, "over_surv_stt")))  

levels(kirc_clinic_numeric$over_surv_stt) <- c("D","L")

ggpairs(kirc_clinic_numeric, columns = cols_numeric, 
        title="Correlation matrix",               
        mapping= aes(colour = over_surv_stt), 
        upper = list(combo = wrap("box_no_facet", alpha=0.1), 
                     continuous = wrap("cor", size = 2, alignPercent = 0.8)),
        lower = list(continuous = wrap("smooth", alpha = 0.3, size=0.2) )) +
        theme(panel.background = element_rect(color = "black", size=0.5, fill="white"),
          panel.grid.major = element_blank()) 
        
```

### 5.1 Run multiple T-tests on over_surv_stt

Transform the data into long format

```{r}
# Put all variables in the same column except `over_surv_stt`, the grouping variable

levels(kirc_clinic_numeric$over_surv_stt) <- c("DECEASED","LIVING")

kirc_clinic_numeric.long <- kirc_clinic_numeric %>%
  pivot_longer(-over_surv_stt, names_to = "variables", values_to = "value")

kirc_clinic_numeric.long <- kirc_clinic_numeric.long[!is.na(kirc_clinic_numeric.long$value), ]

kirc_clinic_numeric.long$value.log <- log2(kirc_clinic_numeric.long$value+1)

kirc_clinic_numeric.long %>% sample_n(6)
```

Group the data by variables and compare over_surv_stt groups

Adjust the p-values and add significance levels

```{r}
stat.test <- kirc_clinic_numeric.long %>%
  group_by(variables) %>%
  t_test(value ~ over_surv_stt) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()
stat.test
```



```{r}
# Create the plot on logscale
myplot <- ggboxplot(
  kirc_clinic_numeric.long, x = "over_surv_stt", y = "value.log",
  fill = "over_surv_stt", palette = "npg", legend = "none", 
  ggtheme = theme_pubr(border = TRUE)
  ) +
  facet_wrap(~variables)
# Add statistical test p-values
stat.test <- stat.test %>% add_xy_position(x = "over_surv_stt")
myplot + stat_pvalue_manual(stat.test, label = "p.adj.signif")
```


```{r}
# Group the data by variables and do a graph for each variable
graphs <- kirc_clinic_numeric.long %>%
  group_by(variables) %>%
  doo(
    ~ggboxplot(
      data =., x = "over_surv_stt", y = "value",
      fill = "over_surv_stt", palette = "npg", legend = "none",
      ggtheme = theme_pubr()
      )  +
      geom_jitter(width = 0.05, alpha = 0.2, color = "orange"), 
    result = "plots"
  )
graphs
```


```{r}

# Add statitistical tests to each corresponding plot
variables <- graphs$variables
for(i in 1:length(variables)){
  graph.i <- graphs$plots[[i]] + 
    labs(title = variables[i]) +
    stat_pvalue_manual(stat.test[i, ], label = "p.adj.signif")
  print(graph.i)
}
```


```{r}

ggplot(kirc_clinic, aes(age, fill= over_surv_stt)) +
  geom_histogram(bins = 15, position = "dodge")
t.test(kirc_clinic$age ~ kirc_clinic$over_surv_stt) 

ggplot(kirc_clinic, aes(year_diagnose, fill= over_surv_stt)) +
  geom_histogram(bins = 15, position = "dodge")
t.test(kirc_clinic$year_diagnose ~ kirc_clinic$over_surv_stt) 

ggplot(kirc_clinic, aes(x=over_surv_stt, y=disease_free_mth)) +
  geom_boxplot(width = .5) +
  geom_jitter(width = 0.05, alpha = 0.2, color = "orange")
t.test(kirc_clinic$disease_free_mth ~ kirc_clinic$over_surv_stt) 

ggplot(kirc_clinic, aes(x=over_surv_stt, y=frac_genome_alter)) +
  geom_boxplot(width = .5) +
  geom_jitter(width = 0.05, alpha = 0.2, color = "orange")
t.test(kirc_clinic$frac_genome_alter ~ kirc_clinic$over_surv_stt)

ggplot(kirc_clinic, aes(x=over_surv_stt, y=long_dim)) +
  geom_boxplot(width = .5) +
  geom_jitter(width = 0.05, alpha = 0.2, color = "orange")
t.test(kirc_clinic$long_dim ~ kirc_clinic$over_surv_stt)

ggplot(kirc_clinic, aes(x=over_surv_stt, y=mutation_cnt)) +
  geom_boxplot(width = .5) +
  geom_jitter(width = 0.05, alpha = 0.2, color = "orange")
t.test(kirc_clinic$mutation_cnt ~ kirc_clinic$over_surv_stt)

ggplot(kirc_clinic, aes(x=over_surv_stt, y=over_surv_mth)) +
  geom_boxplot(width = .5) +
  geom_jitter(width = 0.05, alpha = 0.2, color = "orange")
t.test(kirc_clinic$over_surv_mth ~ kirc_clinic$over_surv_stt)

ggplot(kirc_clinic, aes(x=over_surv_stt, y=short_dim)) +
  geom_boxplot(width = .5) +
  geom_jitter(width = 0.05, alpha = 0.2, color = "orange")
t.test(kirc_clinic$short_dim ~ kirc_clinic$over_surv_stt)

ggplot(kirc_clinic, aes(x=over_surv_stt, y=second_long_dim)) +
  geom_boxplot(width = .5) +
  geom_jitter(width = 0.05, alpha = 0.2, color = "orange")
t.test(kirc_clinic$second_long_dim ~ kirc_clinic$over_surv_stt)

# fazer uma table com as variaveis dependentes, indpendente e p-valores
```

## 4. Categorical variables vs. over_surv_stt

Tabulation and chi-square test

```{r warning=FALSE}
# talvez isso possa sair uma vez que ja tem a mesma analise com tablefit
kirc_clinic %>%
  select_if(is.factor) %>%
  summary() 

t_metas_stg <- table(kirc_clinic$metastasis_stg, kirc_clinic$over_surv_stt, exclude = NULL)
t_metas_stg <- addmargins(round(100*prop.table(t_metas_stg)))
t_metas_stg
chisq.test(x = kirc_clinic$metastasis_stg, y = kirc_clinic$over_surv_stt) 

t_lynph <- table(kirc_clinic$neoplasm_ln_stg, kirc_clinic$over_surv_stt, exclude = NULL)
t_lynph <- addmargins(round(100*prop.table(t_lynph)))
t_lynph
chisq.test(x = kirc_clinic$neoplasm_ln_stg, y = kirc_clinic$over_surv_stt) 

t_neop <- table(kirc_clinic$neoplasm_stg, kirc_clinic$over_surv_stt, exclude = NULL)
t_neop <- addmargins(round(100*prop.table(t_neop)))
t_neop
chisq.test(x = kirc_clinic$neoplasm_stg, y = kirc_clinic$over_surv_stt) 

t_tumor <- table(kirc_clinic$tumor_stg, kirc_clinic$over_surv_stt, exclude = NULL)
t_tumor <- addmargins(round(100*prop.table(t_tumor)))
t_tumor
chisq.test(x = kirc_clinic$tumor_stg, y = kirc_clinic$over_surv_stt) 

t_free <- table(kirc_clinic$disease_free_stt, kirc_clinic$over_surv_stt, exclude = NULL)
t_free <- addmargins(round(100*prop.table(t_free)))
t_free
chisq.test(x = kirc_clinic$disease_free_stt, y = kirc_clinic$over_surv_stt) 

t_prior <- table(kirc_clinic$prior_cancer, kirc_clinic$over_surv_stt, exclude = NULL)
t_prior <- addmargins(round(100*prop.table(t_prior)))
t_prior
chisq.test(x = kirc_clinic$prior_cancer, y = kirc_clinic$over_surv_stt) 

t_neo <- table(kirc_clinic$neoadj_therapy, kirc_clinic$over_surv_stt, exclude = NULL)
t_neo <- addmargins(round(100*prop.table(t_neo)))
t_neo
chisq.test(x = kirc_clinic$neoadj_therapy, y = kirc_clinic$over_surv_stt) 

t_platelet <- table(kirc_clinic$platelet, kirc_clinic$over_surv_stt, exclude = NULL)
t_platelet <- addmargins(round(100*prop.table(t_platelet)))
t_platelet
chisq.test(x = kirc_clinic$platelet, y = kirc_clinic$over_surv_stt)

t_prospect <- table(kirc_clinic$tissue_prospect, kirc_clinic$over_surv_stt, exclude = NULL)
t_prospect <- addmargins(round(100*prop.table(t_prospect)))
t_prospect
chisq.test(x = kirc_clinic$tissue_prospect, y = kirc_clinic$over_surv_stt)

t_race <- table(kirc_clinic$race, kirc_clinic$over_surv_stt, exclude = NULL)
t_race <- addmargins(round(100*prop.table(t_race)))
t_race
chisq.test(x = kirc_clinic$race, y = kirc_clinic$over_surv_stt) 

t_retros <- table(kirc_clinic$tissue_retrospect, kirc_clinic$over_surv_stt, exclude = NULL)
t_retros <- addmargins(round(100*prop.table(t_retros)))
t_retros
chisq.test(x = kirc_clinic$tissue_retrospect, y = kirc_clinic$over_surv_stt)

t_ca <- table(kirc_clinic$serum_ca, kirc_clinic$over_surv_stt, exclude = NULL)
t_ca <- addmargins(round(100*prop.table(t_ca)))
t_ca
chisq.test(x = kirc_clinic$serum_ca, y = kirc_clinic$over_surv_stt) 

t_gender <- table(kirc_clinic$gender, kirc_clinic$over_surv_stt, exclude = NULL)
t_gender <- addmargins(round(100*prop.table(t_gender)))
t_gender
chisq.test(x = kirc_clinic$gender, y = kirc_clinic$over_surv_stt) 

t_site <- table(kirc_clinic$tissue_site, kirc_clinic$over_surv_stt, exclude = NULL)
t_site <- addmargins(round(100*prop.table(t_site)))
t_site
chisq.test(x = kirc_clinic$tissue_site, y = kirc_clinic$over_surv_stt) 

t_neop_st <- table(kirc_clinic$person_neoplasm_stt, kirc_clinic$over_surv_stt, exclude = NULL)
t_neop_st <- addmargins(round(100*prop.table(t_neop_st)))
t_neop_st
chisq.test(x = kirc_clinic$person_neoplasm_stt, y = kirc_clinic$over_surv_stt) 

t_wbc <- table(kirc_clinic$wbc, kirc_clinic$over_surv_stt, exclude = NULL)
t_wbc <- addmargins(round(100*prop.table(t_wbc)))
t_wbc
chisq.test(x = kirc_clinic$wbc, y = kirc_clinic$over_surv_stt) 
```

## 7. FinalFit

summarise variables/factors by a categorical variable

```{r warning=FALSE}
explanatory <- names(kirc_clinic %>%
              select(-over_surv_stt) %>%
              select_if(is.factor))
dependent <-  'over_surv_stt'

table_fit <- kirc_clinic %>%
  summary_factorlist(dependent, explanatory, p=TRUE, add_dependent_label=TRUE)
table_fit
warnings()

#knitr::kable(table_fit, row.names=FALSE, align=c("l", "l", "r", "r", "r"))
```

## Further analysis

- [A logistic regression analysis](3.logistic_regression.md) of each clinical variable weight.

```{r}
sessionInfo()
```

