---
title: "A correlation analysis of clinical variables of TCGA-KIRC patients"
output: 
  html_document: 
    default
  github_document: 
    df_print: paged
  pdf_document:
    latex_engine: xelatex
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all") })     
---

This project contains a pipeline of clinical analysis of the Cancer Genome Atlas Kidney Renal Clear Cell Carcinoma (TCGA-KIRC) data of patients from [Genomic Data Commons Data Portal](https://portal.gdc.cancer.gov/exploration?filters=%7B%22op%22%3A%22and%22%2C%22content%22%3A%5B%7B%22op%22%3A%22in%22%2C%22content%22%3A%7B%22field%22%3A%22cases.project.project_id%22%2C%22value%22%3A%5B%22TCGA-KIRC%22%5D%7D%7D%5D%7D) and [cBioPortal](https://www.cbioportal.org/study/summary?id=kirp_tcga).

Previously, we presented [an exploratory preprocessing analysis](1.preprocessing.md). In this section, we present a correlation analysis with t-test and ANOVA test to investigate significative distinctions between clinical variables according to their vital status. 


```{r, echo = FALSE}
# Code to browse the markdown file with renderized images.
knitr::opts_chunk$set(
  fig.path = "figs/render-"
)
```


```{r message=FALSE, warning=FALSE, paged.print=FALSE, echo = FALSE}
# Set the packages of interest
packages = c("tidyverse","skimr","finalfit")

# if a package is installed, it will be loaded
# otherwise, the missing package(s) will be installed and loaded
package.check <- lapply(packages, FUN = function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x, dependencies = TRUE)
    library(x, character.only = TRUE)
  }
})

suppressMessages(library("tidyverse"))
setwd(".")
```


## 1. Importing data

```{r message=FALSE, warning=FALSE, paged.print=FALSE, echo = FALSE}
kirc_clin <- read_delim("data/kirc_clinical_clean.csv", "\t", 
                            escape_double = FALSE, 
                            trim_ws = TRUE)

```



```{r}
# Update the variables
 # explanatory = c("Age","Metastasis_Stage","Neoplasm_Lymph_Stage","Neoplasm_Stage","Tumor_Stage","Disease_Free_Months",      
 #                 "Disease_Free_Status","Fraction_Genome_Altered","Histologic_Grade","Neoadjuvant_Therapy",
 #                 "Primary_Lymph_Ind3","Overall_Surv_Status","Platelet_count","Sex","Shortest_Dim",            
 #                 "Person_Neoplasm_Status","Patient_Vital_Status")
 # 
 # dependent = "Peak"
 # 
 # dados %>%
 #   summary_factorlist(dependent, explanatory, p = TRUE, na_include = TRUE,
 #                      column = FALSE) -> t
```



## Further analysis

- [A logistic regression analysis](3.logistic_regression.md) of each clinical variable weight.


```{r}
sessionInfo()
```

